---
title: "CSSS508:Tutorial 4"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(checkr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(exercise.checker = checkr::checkr_tutor)
tutorial_options(exercise.timelimit = 10)
vec_1 <- 30:39
vec_2 <- seq(22, 31, length.out = 10)
vec_forward <- c("T", "a", "c", "o", "c", "a", "t")
```


## 1: Coding with (Atomic) Vectors

1. Use `seq()` to create a vector of 10 numbers evenly spaced from 0 to 15. You may want to check the help file for `seq()` using `?seq` to figure out what arguments to use. If you run `?seq` below it will likely open a browser window to display help. Run `?seq` in your R console instead if this causes a problem.
```{r vec_num, exercise=TRUE}

```

```{r vec_num-check, echo=FALSE}
test_1 <- find_call("seq(whatever)", "you need to use `seq()`.")
test_2 <- check_value(match_vector(seq(0,15, length.out=10)), "this is not 10 numbers spaced from 0 to 15.")
USER_CODE %>% test_1 %>% test_2
```


2. Use `:` to create an integer vector the numbers 11 through 20.
```{r vec_int, exercise=TRUE}

```

```{r vec_int-check, echo=FALSE}
test_1 <- find_call("whatever : whatever", "need to use the `:` operator.")
test_2 <- check_value(match_vector(11:20), "this is not the numbers 11 to 20")
USER_CODE %>% test_1 %>% test_2
```


3. Use `LETTERS` and `[ ]` to display the last 10 capital letters. `LETTERS` is a built-in vector of the upper-case letters of the Roman alphabet which you can see below.
```{r vec_cha, exercise=TRUE}

```

```{r vec_cha-check, echo=FALSE}
test_1 <- find_statement("LETTERS", "you need to use `LETTERS`.")
test_2 <- find_call("LETTERS[whatever]", "you need to use `[ ]`.")
test_3 <- check_value(match_vector(LETTERS[17:26]), "this is not the last 10 capital letters.")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```


4. Use `letters`, `[ ]`, and `factor()` to create a factor vector of the first ten lower-case letters.
```{r vec_fac, exercise=TRUE}

```

```{r vec_fac-check, echo=FALSE}
test_1 <- find_value(match_class("factor"), "your output should be a factor.")
test_2 <- check_argument("factor(grab_this)", test=match_vector(letters[1:10]), "this is not the first 10 lower-case letters.")
test_3 <- find_statement("letters", "you need to use `letters`.")
test_4 <- find_call("letters[whatever]", "you need to use `[ ]`.")
USER_CODE %>% test_1 %>% test_2 %>% test_3 %>% test_4
```


5. Use `c()` to combine `vec_1` and `vec_2`. You can enter `vec_1` or `vec_2` below and run them if you want to see what they are.
```{r vec_12, exercise=TRUE}


```

```{r vec_12-check, echo=FALSE}
test_1 <- find_call("c(whatever)", "you need to use `c()`.")
test_2 <- check_value(match_vector(c(30:39, seq(22, 31, length.out = 10))), "this is not the concatenation of `vec_1` and `vec_2`.")
USER_CODE %>% test_1 %>% test_2
```


6. Use `c()` and `[ ]` to combine the first 4 elements of `vec_1` with the last 4 elements of `vec_2`.
```{r vec_12_2, exercise=TRUE}


```

```{r vec_12_2-check, echo=FALSE}
test_1 <- find_call("c(whatever)", "you need to use c().")
test_2 <- find_call("vec_1[whatever]", "you need to use `[ ]` with `vec_1`.")
test_3 <- find_call("vec_2[whatever]", "you need to use `[ ]` with `vec_2`.")
test_4 <- check_value(match_vector(c(vec_1[1:4], vec_2[1:4])), "this is not the concatenation of the first 4 elements each of `vec_1` and `vec_2`.")
USER_CODE %>% test_1 %>% test_2 %>% test_3 %>% test_4
```


7. Use `rev()` to reverse the order of `vec_forward`. You can enter `vec_forward` below and run the code if you want to see what it is.
```{r vec_reverse, exercise=TRUE}


```

```{r vec_reverse-check, echo=FALSE}
test_1 <- find_call("rev(whatever)", "you need to use `rev()`.")
test_2 <- check_value(match_vector(rev(vec_forward)), "this is not the reverse of `vec_forward`.")
USER_CODE  %>% test_1 %>% test_2
```

# 2: Coding with matrices
