---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(checkr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(exercise.checker = checkr::checkr_tutor)
```


# 1: Coding with (Atomic) Vectors

1. Use `seq()` to create a vector named `vec_num` of 10 numbers evenly spaced from 0 to 15.
```{r vec_num, exercise=TRUE, echo=FALSE}

```

```{r vec_num-check, echo=FALSE}
test_1 <- find_assignment("vec_num", "the vector should be named vec_num.")
test_2 <- find_call("seq(whatever)", "you need to use seq().")
test_3 <- check_value(match_vector(seq(0,15, length.out=10)), "this is not 10 numbers spaced from 0 to 15.")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```


2. Use `:` to create an integer vector named `vec_int` of the numbers 11 through 20.
```{r vec_int, exercise=TRUE, echo=FALSE}

```

```{r vec_int-check, echo=FALSE}
test_1 <- find_assignment("vec_int", "the vector should be named vec_int.")
test_2 <- find_call("whatever : whatever", "need to use the `:` operator.")
test_3 <- check_value(match_vector(11:20), "this is not the numbers 11 to 20")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```

3. Use `LETTERS` and `[ ]` to create a vector named `vec_cha` of the last 10 capital letters. `LETTERS` is a built-in vector of the upper-case letters of the Roman alphabet.
```{r vec_cha, exercise=TRUE, echo=FALSE}

```

```{r vec_cha-check, echo=FALSE}
test_1 <- find_assignment("vec_cha", "the vector should be named vec_cha.")
test_2 <- check_value(match_vector(LETTERS[17:26]), "this is not the last 10 capital letters.")
test_3 <- find_statement("LETTERS", "you need to use `LETTERS`.")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```

4. Use `letters` and `[ ]` to create a factor variable using the first ten lower-case letters.
```{r vec_fac, exercise=TRUE, echo=FALSE}

```

```{r vec_fac-check, echo=FALSE}
test_1 <- find_assignment("vec_fac", "the vector should be named vec_fac.")
test_2 <- find_value(match_class("factor"), "vec_fac should be a factor.")
# test_3 <- check_value(match_vector(factor(letters[1:10]), same_order = F), "this is not the first 10 lower-case letters.")
test_3 <- check_argument("factor(grab_this)", test=match_vector(letters[1:10]), "this is not the first 10 lower-case letters.")
test_4 <- find_statement("letters", "you need to use `letters`.")
USER_CODE %>% test_1 %>% test_2 %>% test_3 %>% test_4
```

5. Use `c()` to combine `vec_1` and `vec_2` into a vector named `vec_12`. Do not edit the first line where `vec_1` and `vec_2` are created.
```{r vec_12-setup}
vec_1 <- 30:39
vec_2 <- seq(22, 31, length.out = 10)
```

```{r vec_12, exercise=TRUE, echo=FALSE, exercise.setup="prepare-problem_1"}


```

```{r vec_12-check, echo=FALSE}
test_1 <- find_assignment("vec_12", "the vector should be named vec_12.")
test_2 <- find_call("c(whatever)", "you need to use c().")
test_3 <- check_value(match_vector(c(30:39, seq(22, 31, length.out = 10))), "this is not the concatenation of `vec_1` and `vec_2`.")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```

6. Use "c()" and "[ ]" to combine the first 4 elements of "vec_num" with the last 4 elements of "vec_int" to create "vec_ni".
vec_ni <- NULL

7. Use "rev()" to reverse the order of "vec_fac".
fac_vec <- NULL
