---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(checkr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(exercise.checker = checkr::checkr_tutor)
```


## 1: Coding with (Atomic) Vectors

1. Use `seq()` to create a vector named `vec_num` of 10 numbers evenly spaced from 0 to 15.
```{r vec_num, exercise=TRUE, echo=FALSE}

```

```{r vec_num-check, echo=FALSE}
test_1 <- find_assignment("vec_num", "the vector should be named vec_num.")
test_2 <- find_call("seq(whatever)", "you need to use seq().")
test_3 <- check_value(match_vector(seq(0,15, length.out=10)), "this is not 10 numbers spaced from 0 to 15.")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```


2. Use ":" to create an integer vector named `vec_int` of the numbers 11 through 20.
```{r vec_int, exercise=TRUE, echo=FALSE}

```

```{r vec_int-check, echo=FALSE}
test_1 <- find_assignment("vec_int", "the vector should be named vec_int.")
test_2 <- find_call("whatever : whatever", "need to use the : operator.")
test_3 <- check_value(match_vector(11:20), "this is not the numbers 11 to 20")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```

3. Use "LETTERS" and "[ ]" to create a vector of the last 10 capital letters.
```{r vec_cha, exercise=TRUE, echo=FALSE}

```

```{r vec_cha-check, echo=FALSE}
test_1 <- find_assignment("vec_cha", "the vector should be named vec_cha.")
test_2 <- check_value(match_vector(LETTERS[17:26]), "this is not the last 10 capital letters.")
test_3 <- find_statement("LETTERS", "you need to use LETTERS.")
USER_CODE %>% test_1 %>% test_2 %>% test_3
```

4. Use letters and "[ ]" to create a factor variable using the first ten lower case letters.
vec_fac <- NULL

# 5) Use "c()" to combine "vec_cha" and "vec_fac" into "vec_let". Do not convert it to a factor!
vec_let <- NULL

# 6) Use "c()" and "[ ]" to combine the first 4 elements of "vec_num" with the last
# 4 elements of "vec_int" to create "vec_ni".
vec_ni <- NULL

# 7) Use "rev()" to reverse the order of "vec_fac".
fac_vec <- NULL
