---
title: "CSSS508, Week 7"
subtitle: "Vectorization and Functions"
author: "Chuck Lanfear"
date: "May 13, 2020<br>Updated: `r gsub(' 0', ' ', format(Sys.Date(), format='%b %d, %Y'))`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["center","top"]
---

```{r setup, include=FALSE, purl=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "##")
```

class: inverse
# A Quick Aside

---
# Visualize the Goal First

Before you can write effective code, you need to know *exactly* what you want that code to produce.

--

* Do I want a single value? A vector? List?
* Do I want one observation per person? Person-year? Year?

--

Most programming problems can be reduced to having an unclear idea of your end **goal** (or your beginning state).

--

If you know what you *have* (the data structure) and what you *want*, the intermediate steps are usually obvious.

--

When in doubt, *sketch* the beginning state and the intended end state. Then consider what translates the former into the latter in the least complicated way.

--

If that seems complex, break it into more steps.

---
class: inverse
# Vectorization

---
# Example from Last Week

Remember when we tried find the mean for each variable in the `swiss` data?

The best solution is to just use `colMeans()` without even thinking about pre-allocation or `for()` loops:
```{r}
colMeans(swiss)
```

---
# Vectorization Avoids Loops

Loops are very powerful and applicable in almost any situation.

--

They are also slow and require writing more code than vectorized commands.

--

Whenever possible, use existing vectorized commands like `colMeans()` or `dplyr` functions.

--

Sometimes no functions exist to do what you need, so you'll be tempted to write a loop.

This makes sense on a *fast, one-time operation, on small data*.

--

If your data are large or you're going to do it repeatedly, however, consider *writing your own functions*!

---
class: inverse
# Writing Functions

---
# Examples of Existing Functions

* `mean()`:
    + Input: a vector
    + Output: a single number

--

* `dplyr::filter()`:
    + Input: a data frame, logical conditions
    + Output: a data frame with rows removed using those conditions

--

* `readr::read_csv()`:
    + Input: a file path, optionally variable names or types
    + Output: a data frame containing info read in from file

---
# Why Write Your Own Functions?

Functions can encapsulate repeated actions such as:

--

* Given a vector, compute some special summary stats
* Given a vector and definition of "invalid" values, replace with `NA`
* Templates for favorite `ggplot`s used in reports

--

Advanced function applications (not covered in this class):

* Parallel processing
* Generating *other* functions
* Making custom packages containing your functions

---
# Simple Function

Let's look at a function that takes a vector as input and outputs a named vector of the first and last elements:

```{r}
first_and_last <- function(x) {
    first <- x[1]
    last  <- x[length(x)]
    return(c("first" = first, "last" = last))
}
```

--

Test it out:

```{r}
first_and_last(c(4, 3, 1, 8))
```

---
# Testing `first_and_last`

What if I give `first_and_last()` a vector of length 1?

```{r}
first_and_last(7)
```

--

Of length 0?

```{r}
first_and_last(numeric(0))
```

--

Maybe we want it to be a little smarter.

---
# Checking Inputs

Let's make sure we get an error message when the vector is too small:

```{r}
smarter_first_and_last <- function(x) {
    if(length(x) == 0L) { # specify integers with L
        stop("The input has no length!") #<<
    } else {
        first <- x[1]
        last  <- x[length(x)]
        return(c("first" = first, "last" = last))        
    }
}
```

.footnote[`stop()` ceases running the function and prints the text inside as an error message.]
---
# Testing Smarter Function

```{r, error=TRUE}
smarter_first_and_last(numeric(0))
smarter_first_and_last(c(4, 3, 1, 8))
```

---
# Cracking Open Functions

If you type a function name without any parentheses or arguments, you can see its contents:

```{r}
smarter_first_and_last
```

You can also put your cursor over a function in your syntax and hit `F2`.

---
# Anatomy of a Function

.small[
```{r, eval=FALSE}
NAME <- function(ARGUMENT1, ARGUMENT2=DEFAULT){
  BODY
  return(OUTPUT)
}
```
]

* **Name**: What you assign the function to so you can use it later
    + You can have "anonymous" (no-name) functions
--
* **Arguments** (aka inputs, parameters): things the user passes to the function that affect how it works
    + e.g. `x` or `na.rm` in `my_new_func <- function(x, na.rm = FALSE) {...}`
    + `na.rm = FALSE` is example of setting a default value: if user doesn't say what `na.rm` is, it'll be `FALSE`
    + `x`, `na.rm` values won't exist in R outside of the function
--
* **Body**: The actual operations inside the function.
--

* **Return Value**: The output inside `return()`. Could be a vector, list, data frame, another function, or even nothing
    + If unspecified, will be the last thing calculated (maybe not what you want?)
    
---
# Example: Reporting Quantiles

Maybe you want to know more detailed quantile information than `summary()` gives you and with interpretable names.

Here's a starting point:

.smallish[
```{r}
quantile_report <- function(x, na.rm = FALSE) {
    quants <- quantile(x, na.rm = na.rm, 
       probs = c(0.01, 0.05, 0.10, 0.25, 0.5, 0.75, 0.90, 0.95, 0.99))
    names(quants) <- c("Bottom 1%", "Bottom 5%", "Bottom 10%", "Bottom 25%",
                       "Median", "Top 25%", "Top 10%", "Top 5%", "Top 1%")
    return(quants)
}
quantile_report(rnorm(10000))
```
]

---
class: inverse
# An Aside on Apply functions

---
### Don't Loop, `apply()` Yourself Instead

Writing loops is challenging, particularly for new coders.

Loops also require writing a lot of code and are hard to troubleshoot.

--

But loops aren't the only way to iterate in R.

--

Like a loop, `apply` functions iterate over elements of objects, except:

* They don't need preallocation--you can directly assign the output.
* They *must use a function*

*Nearly anything you can do with a loop can be done more easily with the `apply` family of functions*

---
# `lapply()`: List + Functions

`lapply()` is used to **apply** a function over a **l**ist of any kind (e.g. a data frame) and return a list. This is a lot easier than preparing a `for()` loop!

```{r, eval=FALSE}
lapply(swiss, FUN = quantile_report)
```

.small[
```{r, echo=FALSE}
lapply(swiss, FUN = quantile_report)[1:3]
```
]

---
#  `sapply()`: Simple `lapply()` 

A downside to `lapply()` is that lists are hard to work with. `sapply()` **s**implifies the output by making each element a column in a matrix... usually:

.small[
```{r}
sapply(swiss, FUN = quantile_report)
```
]

---
# `apply()`

There is also `apply()` which works over matrices or data frames. You can apply the function to each row (`MARGIN = 1`) or column (`MARGIN = 2`).

.small[
```{r}
apply(swiss, MARGIN = 2, FUN = quantile_report)
```
]
---
# Data Loading with Loop

Remember the loop for loading data files from last week?

```{r, message = FALSE, warnings = FALSE}
library(dplyr); library(readr)
file_list  <- list.files("./example_data/")
file_paths <- paste0("./example_data/", file_list)
data_names <- stringr::str_remove(file_list, ".csv")
data_list  <- vector("list", length(file_list))
names(data_list) <- data_names
for (i in seq_along(file_list)){
  data_list[[ data_names[i] ]] <- read_csv(file_paths[i])
} 
complete_data <- bind_rows(data_list)
head(complete_data, 3)
```

---
# Data Loading with `lapply()`

Another way to load these files would be to... `lapply()` over the file names then bind the rows together. Faster and easier!

```{r, warning=FALSE, message=FALSE}
complete_data <- lapply(file_paths, read_csv) %>%
  bind_rows()
head(complete_data, 3)
```

---
# Data Loading with `vroom`

The fastest and easiest way is to use a fully vectorized data loading function, like `vroom::vroom()`!

```{r, warning=FALSE, message=FALSE}
library(vroom)
complete_data <- vroom(file_paths)
head(complete_data, 3)
```

Just give `vroom()` a vector of file locations and it determines their delimiter, loads them all (crazy fast), and binds them into one dataframe.

---
## From Loop to `apply()`

Converting code in a loop to an `apply` function is straightforward:

1. What you iterate over in the loop (e.g. `seq_along(x)`) becomes the first input.

2. The body of the loop becomes a function.

   * This function should take only the iterator index (e.g. `i`) as an input.
   
3. Assign the output to what your loop stored values in.

---
class: inverse
# Back to Making functions!

---
# Example: Discretizing Continuous Data

Maybe you often want to bucket variables in your data into groups based on quantiles:

| Person | Income | Income Bucket |
|:------:|-------:|--------------:|
|    1   |   8000 |             1 |
|    2   | 103000 |             3 |
|    3   |  12000 |             1 |
|    4   |  52000 |             2 |
|    5   | 150000 |             3 |
|    6   |  45000 |             2 |

---
# Bucketing Function

There's already a function in R called `cut()` that does this, but you need to tell it cutpoints or the number of buckets.

Let's make a convenience function that calls `cut()` using quantiles for splitting and returns an integer:

```{r}
bucket <- function(x, quants = c(0.333, 0.667)) {
    # set low extreme, quantile points, high extreme
    new_breaks <- c(min(x)-1, quantile(x, probs = quants), max(x)+1)
    # labels = FALSE will return integer codes instead of ranges
    return(cut(x, breaks = new_breaks, labels = FALSE))
}
```

---
# Trying Out `bucket()`

.smallish[
```{r, fig.height=4, dev="svg"}
random_data <- rnorm(100)
bucketed_random_data <- bucket(random_data, 
                          quants = c(0.05, 0.25, 0.5, 0.75, 0.95))
plot(x = bucketed_random_data, y = random_data, main = "Buckets and values")
```
]

---
# Example: Removing Bad Data

Let's say we have data where impossible values occur:

.smallish[
```{r}
(school_data <- 
  data.frame(school = letters[1:10],
  pr_passing_exam=c(0.78, 0.55, 0.91,   -1, 0.88, 0.81, 0.90, 0.76,   99, 99),
  pr_free_lunch = c(0.33,   99, 0.25, 0.05, 0.12, 0.09, 0.22,  -13, 0.21, 99)))
```
]

---
# Function to Remove Extreme Values

Goal:

* Input: a vector `x`, cutoff for `low`, cutoff for `high`
* Output: a vector with `NA` in the extreme places

--

```{r}
remove_extremes <- function(x, low, high) {
    x_no_low <- ifelse(x < low, NA, x)
    x_no_low_no_high <- ifelse(x_no_low > high, NA, x)
    return(x_no_low_no_high)
}
remove_extremes(school_data$pr_passing_exam, low = 0, high = 1)
```

---
# `dplyr::mutate_at()`

`dplyr` functions `mutate_at()` and `summarize_at()` take an argument `funs()`. This will apply our function to every variable (besides `school`) to update the columns in `school_data`:

.smallish[
```{r, message=FALSE, warning=FALSE}
library(dplyr)
school_data %>%
   mutate_at(vars(-school), ~ remove_extremes(x = ., low = 0, high = 1))
```
]

---
# Standard and Non-Standard Evaluation

`dplyr` uses what is called **non-standard evaluation** that lets you refer to "naked" variables (no quotes around them) like `school`.

`dplyr` verbs (like `mutate()`) recently started supporting *standard evaluation* allowing you to use quoted object names as well. This makes programming with `dplyr` easier.

```{r}
swiss %>%
  select("Fertility", "Catholic") %>%
  head(2)
```

---
# Anonymous Functions in `dplyr`

You can skip naming your function in `dplyr` if you won't use it again. Code below will return the mean divided by the standard deviation for each variable in `swiss`:

```{r}
swiss %>%
    summarize_all( ~ mean(., na.rm = TRUE) / sd(., na.rm = TRUE))
```

---
# Anonymous `lapply()`

Like with `dplyr`, you can use anonymous functions in `lapply()`, but a difference is you'll need to have the `function()` part at the beginning:

.smallish[
```{r, eval=FALSE}
lapply(swiss, function(x) mean(x, na.rm = TRUE) / sd(x, na.rm = TRUE))
```
]

```{r, echo=FALSE}
lapply(swiss, function(x) mean(x, na.rm = TRUE) / sd(x, na.rm = TRUE))[1:5]
```

---
# Example: `ggplot2` Templates

Let's say you have a particular way you like your charts:

```{r, warning=FALSE, message=FALSE, eval=FALSE}
library(gapminder); library(ggplot2)
ggplot(gapminder %>% filter(country == "Afghanistan"),
       aes(x = year, y = pop / 1000000)) +
       geom_line(color = "firebrick") +
       xlab(NULL) + ylab("Population (millions)") +
       ggtitle("Population of Afghanistan since 1952") +
       theme_minimal() + 
       theme(plot.title = element_text(hjust = 0, size = 20))
```

--

* How could we make this flexible for any country?
--

* How could we make this flexible for any `gapminder` variable?

---
# Example of Desired Chart

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 4, dev="svg"}
library(gapminder); library(ggplot2)
ggplot(gapminder %>% filter(country == "Afghanistan"),
       aes(x = year, y = pop / 1000000)) +
    geom_line(color = "firebrick") +
    xlab(NULL) + ylab("Population (millions)") +
    ggtitle("Population of Afghanistan since 1952") +
    theme_minimal() + theme(plot.title = element_text(hjust = 0, size = 20))
```

---
# Another Example

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 4, dev="svg"}
ggplot(gapminder %>% filter(country == "Peru"),
       aes(x = year, y = lifeExp)) +
    geom_line(color = "firebrick") +
    xlab(NULL) + ylab("Life expectancy") +
    ggtitle("Life expectancy in Peru since 1952") +
    theme_minimal() + theme(plot.title = element_text(hjust = 0, size = 20))
```

---
# Making Country Flexible

We can have the user input a character string for `cntry` as an argument to the function to get subsetting and the title right:
```{r}
gapminder_lifeplot <- function(cntry) {
    ggplot(gapminder %>% filter(country == cntry), #<<
       aes(x = year, y = lifeExp)) +
    geom_line(color = "firebrick") +
    xlab(NULL) + ylab("Life expectancy") + theme_minimal() + 
    ggtitle(paste0("Life expectancy in ", cntry, " since 1952")) + #<<
    theme(plot.title = element_text(hjust = 0, size = 20))
}
```

What `cntry` does:

* `filter()` to the specific value of `cntry`

* Add text value of `cntry` in `ggtitle()`

---
# Testing Plot Function

```{r, warning=FALSE, message=FALSE, echo=TRUE, fig.height = 4, dev="svg"}
gapminder_lifeplot(cntry = "Turkey")
```

---
# Making `y` Value Flexible

Now let's allow the user to say which variable they want on the y-axis. How we can get the right labels for the axis and title? We can use a named character vector to serve as a "lookup table" inside the function:

.smallish[
```{r}
y_axis_label <- c("lifeExp" = "Life expectancy",
                  "pop" = "Population (millions)",
                  "gdpPercap" = "GDP per capita, USD")
title_text <- c("lifeExp" = "Life expectancy in ",
                "pop" = "Population of ",
                "gdpPercap" = "GDP per capita in ")
# example use:
y_axis_label["pop"]
title_text["pop"]
```
]

---
# `aes_string()`

`ggplot()` is usually looking for "naked" variables, but we can tell it to take them as quoted strings (standard evaluation) using `aes_string()` instead of `aes()`, which is handy when making functions:
```{r}
gapminder_plot <- function(cntry, yvar) {
    y_axis_label <- c("lifeExp" = "Life expectancy",
                      "pop" = "Population (millions)",
                      "gdpPercap" = "GDP per capita, USD")[yvar] #<<
    title_text   <- c("lifeExp" = "Life expectancy in ",
                      "pop" = "Population of ",
                      "gdpPercap" = "GDP per capita in ")[yvar] #<<
    ggplot(gapminder %>% filter(country == cntry) %>% #<<
             mutate(pop = pop / 1000000),
           aes_string(x = "year", y = yvar)) + #<<
      geom_line(color = "firebrick") + 
      ggtitle(paste0(title_text, cntry, " since 1952")) +  #<<
      xlab(NULL) + ylab(y_axis_label) + theme_minimal() +
      theme(plot.title = element_text(hjust = 0, size = 20))
}
```

---
# Testing `gapminder_plot()`

```{r, warning=FALSE, message=FALSE, echo=TRUE, fig.height = 4, dev="svg"}
gapminder_plot(cntry = "Turkey", yvar = "pop")
```

---
# Debugging

Something not working as hoped? Try using `debug()` on a function, which will show you the world as perceived from inside the function:

```{r, eval=FALSE}
debug(gapminder_plot)
```

Then when you've fixed your problem, use `undebug()` so that you won't go into debug mode every time you run it:

```{r, eval=FALSE}
undebug(gapminder_plot)
```

---
# Overview: The Process

Data processing can be very complicated, with many valid ways of accomplishing it.

I believe the best general approach is the following:
--

1. Look carefully at the **starting data** to figure out what you can get from them.
--

2. Determine *precisely* what you want the **end product** to look like.
--

3. Identify individual steps needed to go from Step 1 to Step 2.
--

4. Make each discrete step its own set of functions or function calls.
   + If any step is confusing or complicated, **break it into more steps**.
--

5. Complete each step *separately and in order*.
   + Do not continue until a step is producing what you need for the next step.
   + **Do not worry about combining steps for efficiency until everything works**.

--

Once finished, if you need to do this again, *convert the prior steps into functions*!

---
# Bonus Function

My lectures are rendered with a function!

.smallish[
```{r, eval=FALSE}
render_and_print_slides <- function(week){
    week_dir     <- paste0(getwd(), "/Lectures/", "Week", week, "/")
    current_rmd  <- paste0(week_dir, stringr::str_subset(list.files(week_dir),
                                                     "^CSSS508_Week.*\\.Rmd$"))
    rmarkdown::render(current_rmd, encoding = "UTF-8")
    current_html <- stringr::str_replace(current_rmd,  "\\.Rmd",  "\\.html")
    new_pdf_file <- stringr::str_replace(current_html, "\\.html", "\\.pdf")
    new_r_script <- stringr::str_replace(current_html, "\\.html", "\\.R")
    message("Slides rendered, waiting 5 seconds.")
    Sys.sleep(5)
    message("Purling slides.")
    knitr::purl(input = current_rmd, output = new_r_script, documentation = 0)
    message("Printing from Chrome.")
    pagedown::chrome_print(current_html, format="pdf")
    message(paste0("Printing complete at ", week_dir))
}
```
]

I give it a numeric week and it (1) finds the lecture `.Rmd`, (2) knits the slides, (3) creates a `.R` file, (4) then opens the slides in Chrome and prints a PDF.

---
class: inverse
# Homework

[Download](https://s3.amazonaws.com/pronto-data/open_data_year_one.zip) and analyze data from the first year of Seattle's Pronto! bike sharing program.

Using the provided template, you will write: 

1. A loop (or `lapply()`) to read in the data from multiple files.
2. Functions to clean up the data 
3. A function to visualize ridership over the first year.

There is some string processing needed—much of which you have already seen or can probably Google—but *some will come in the next lecture*. I give suggestions in the template, but I can cover string processing in detail in lab if needed before the homework is due.

## PART 1 DUE: Next week
## PART 2 DUE: In two weeks
